%-----------------------------------------------
%  Engineer's Thesis
%  Warsaw University of Technology
%-----------------------------------------------

\documentclass[
    bindingoffset=5mm,
    footnoteindent=3mm,
    hyphenation=true
]{src/wut-thesis}

\graphicspath{{tex/img/}}
\addbibresource{bibliografia.bib}

%-------------------------------------------------------------
% Configuration
%-------------------------------------------------------------
\facultyeiti
\EngineerThesis
\langeng

\begin{document}

%------------------
% Title Page
%------------------
\institute{Control and Computation Engineering}
\major{Computer Science}
\specialization{Software Engineering}
\title{
    Web/desktop application (TypeScript) for visualizing and composing dependency graphs between reusable software components
}
\engtitle{
    Web/desktop application (TypeScript) for visualizing and composing dependency graphs between reusable software components
}
\author{Binh Vuong Le Duc}
\album{330097}
\supervisor{mgr inż. Sławomir Niespodziany}
\date{\the\year}
\maketitle

%-------------------------
% Abstract
%-------------------------
\cleardoublepage
\abstract
Modern embedded systems development increasingly relies on modular software architectures to manage complexity and enhance reusability. The \texttt{diff} framework, a custom-built dependency injection system for resource-constrained C++ environments, enables dynamic system configuration through JSON-based topology files. However, as the number of interconnected components grows, manual maintenance of these configuration files becomes error-prone and lacks intuitive structural oversight.

This thesis presents the design and implementation of \texttt{diff-forge}, a desktop application developed with TypeScript, React, and Electron, specifically tailored for the visual composition of software topologies within the \texttt{diff} ecosystem. The proposed solution provides a graphical canvas for managing component instances and their dependencies, complemented by a real-time validation engine for cycle and orphan detection. By bridging the gap between abstract topology definitions and visual representation, \texttt{diff-forge} aims to improve developer productivity and reduce configuration errors in the assembly of complex embedded software systems.

\keywords
TypeScript, Electron, Dependency Injection, Visual Programming, Embedded Systems

\pagestyle{plain}

%-------------------------
% Table of Contents
%-------------------------
\cleardoublepage
\tableofcontents

%-------------------------
% Chapters
%-------------------------
\cleardoublepage
\pagestyle{headings}

\input{tex/1-introduction}
\input{tex/2-background}
\input{tex/3-framework}
\input{tex/4-design}
\input{tex/5-implementation}
\input{tex/6-evaluation}
\input{tex/7-conclusion}

%-------------------------
% Bibliography
%-------------------------
\cleardoublepage
\printbibliography
\clearpage

%-------------------------
% Symbols and Abbreviations
%-------------------------
\acronymlist
\acronym{EiTI}{Faculty of Electronics and Information Technology}
\acronym{WUT}{Warsaw University of Technology}
\acronym{DI}{Dependency Injection}
\acronym{IAiIS}{Institute of Control and Computation Engineering}
\acronym{CBSE}{Component-Based Software Engineering}
\acronym{GUI}{Graphical User Interface}
\vspace{0.8cm}

%-------------------------
% Lists: Figures, Tables, Appendices
%-------------------------
\pagestyle{plain}

\listoffigurestoc
\vspace{1cm}
\listoftablestoc
\vspace{1cm}
\listofappendicestoc

%-------------------------
% Appendices
%-------------------------
\captionsetup[figure]{list=no}
\captionsetup[table]{list=no}

\clearpage
\appendix{User Manual}
Instructions for installing and running \texttt{diff-forge}.

\clearpage
\appendix{Configuration Schema}
The JSON schema used for component metadata.

\end{document}
